New approaches to teachers' experience of stress: Do heart rate measurements with fitness trackers provide an efficient, inexpensive, and robust measurement method?

### 2022-09-15 02_paper/02_study/02_fitbit paper/fitbit_paper.Rmd
update fitbit

---
title             : "A Walk In The Classroom - Using Fitbit Charge 4 to measure expert and novice teachers’ heart rate and step counts in a laboratory study"
shorttitle        : "Measuring novice and expert teachers' heart rate"

author: 
  - name          : "Mandy Klatt"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Dittrichring 5-7, 04109 Leipzig"
    email         : "mandy.klatt@uni-leipzig.de"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Peer Keßler"
    affiliation   : "1"
    role:
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Gregor Kachel"
    affiliation   : "1,2"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
  - name          : "Anne Deiglmayr"
    affiliation   : "1"
    role:
      - "Supervision"

@@

```
# Introduction
Especially the first five years are a crucial time the teaching profession, whether the development from novice to expert is experienced as successful and as an increase in competence or leads to the fact that between 40 and 50 percent of the beginning teachers left the profession [@ingersoll2003wrong; @prilleltensky2016teacher]. One reason among many why teachers end their careers is the demanding workload and resulting stress in the teaching profession [@struyven2014teachers]. In her study with 400 secondary teachers, @fisher2011factors showed that burnout scores were significantly different between novice and experienced teachers, whereby novices had higher burnout. Furthermore, she found that low years of experience was a statistically significant predictor for stress in the teaching profession, along with job satisfaction and burnout.

XXX ÜBERGANG VON STRESS ZU HEART RATE XXX

The heart rate is considered an indicator of physical or emotional stress. Increased workload is accompanied by increased heart rate. Therefore, heart rate can be considered the most important psycho physiological parameter [@sachs2014physiologische].

XXX ÜBERGANG VON HEART RATE ZU SMARTWATCHES - VORTEILE XXX low-cost, non-invasive, wrist-worn devices

There is a gap in research on whether heart rate measurements using fitness watches are appropriate to detect differences between experienced and inexperienced teachers in a laboratory setting.

>>>>

---
title             : "New approaches to teachers' experience of stress: Do heart rate measurements with fitness trackers provide an efficient, inexpensive, and robust measurement method?"
shorttitle        : "Measuring novice and expert teachers' heart rate"

author: 
  - name          : "Mandy Klatt"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Dittrichring 5-7, 04109 Leipzig"
    email         : "mandy.klatt@uni-leipzig.de"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Peer Keßler"
    affiliation   : "1"
    role:
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Gregor Kachel"
    affiliation   : "1,2"
    role:
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Christin Lotz"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Anne Deiglmayr"
    affiliation   : "1"
    role:
      - "Supervision"

@@

```
# Introduction

Prospective teachers experience the first five years as a particular professional challenge, as they are associated with a high workload (e.g., class size, poor student behavior) and resulting stress in the teaching profession [@struyven2014teachers].

Psychosomatic stress symptoms, such as increasing heart rate, result from an interaction of situational stressors and person-specific available resources ["transactional stress model"; @lazarus1987transactional; @obbarius2021modified]. If teachers are exposed to a stressful teaching-learning environment for a long time and do not have the sufficient resources and coping strategies, this can lead to burnout.

Fisher (2011) showed in her study of 400 secondary school teachers that burnout scores differed significantly between novice and experienced teachers, with novices having higher burnout scores. In addition, she found that a low number of years of professional experience was a statistically significant predictor of stress in the teaching profession, along with job satisfaction and burnout.

In addition to self-reported indicators of stress (e.g., emotional exhaustion), psychophysiological indicators (e.g., heart rate) are becoming increasingly important in research. Heart rate as the number of heart contractions per minute [@Löllgen2015] is an important indicator of physical or emotional stress. Increased workload is associated with increased heart rate [@sachs2014physiologische]. Arguably, physiological data provide a more objective and non-reactive measure of stress in the classroom (Runge et al., 2020). However, in an educational context, this requires the use of low-cost, non-invasive instruments. Fitness trackers worn on the wrist could fill this research gap [@ferguson2015validity].

### 2022-09-15 02_paper/02_study/02_fitbit paper/fitbit_paper.tex
update fitbit

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Walk In The Classroom - Using Fitbit Charge 4 to measure expert and novice teachers' heart rate and step counts in a laboratory study},
  pdfauthor={Mandy Klatt1, Peer Keßler1, Gregor Kachel1,2, \& Anne Deiglmayr1},
  pdflang={en-EN},
  pdfkeywords={heart rate; photoplethysmography; wearable electronic device; expertise differences},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A Walk In The Classroom - Using Fitbit Charge 4 to measure expert and novice teachers' heart rate and step counts in a laboratory study}
\author{Mandy Klatt\textsuperscript{1}, Peer Keßler\textsuperscript{1}, Gregor Kachel\textsuperscript{1,2}, \& Anne Deiglmayr\textsuperscript{1}}
\date{}

@@

The authors made the following contributions. Mandy Klatt: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Peer Keßler: Writing - Original Draft Preparation, Writing - Review \& Editing; Gregor Kachel: Conceptualization, Writing - Review \& Editing; Anne Deiglmayr: Supervision.

@@

Especially the first five years are a crucial time the teaching profession, whether the development from novice to expert is experienced as successful and as an increase in competence or leads to the fact that between 40 and 50 percent of the beginning teachers left the profession (Ingersoll \& Smith, 2003; Prilleltensky, Neff, \& Bessell, 2016). One reason among many why teachers end their careers is the demanding workload and resulting stress in the teaching profession (Struyven \& Vanthournout, 2014). In her study with 400 secondary teachers, Fisher (2011) showed that burnout scores were significantly different between novice and experienced teachers, whereby novices had higher burnout. Furthermore, she found that low years of experience was a statistically significant predictor for stress in the teaching profession, along with job satisfaction and burnout.

XXX ÜBERGANG VON STRESS ZU HEART RATE XXX

The heart rate is considered an indicator of physical or emotional stress. Increased workload is accompanied by increased heart rate. Therefore, heart rate can be considered the most important psycho physiological parameter (Sachs, 2014).

XXX ÜBERGANG VON HEART RATE ZU SMARTWATCHES - VORTEILE XXX low-cost, non-invasive, wrist-worn devices

There is a gap in research on whether heart rate measurements using fitness watches are appropriate to detect differences between experienced and inexperienced teachers in a laboratory setting.

@@

\leavevmode\vadjust pre{\hypertarget{ref-fisher2011factors}{}}%
Fisher, M. H. (2011). Factors influencing stress, burnout, and retention of secondary teachers. \emph{Current Issues in Education}, \emph{14}(1).

@@

\leavevmode\vadjust pre{\hypertarget{ref-ingersoll2003wrong}{}}%
Ingersoll, R. M., \& Smith, T. M. (2003). The wrong solution to the teacher shortage. \emph{Educational Leadership}, \emph{60}(8), 30--33.

@@

\leavevmode\vadjust pre{\hypertarget{ref-prilleltensky2016teacher}{}}%
Prilleltensky, I., Neff, M., \& Bessell, A. (2016). Teacher stress: What it is, why it's important, how it can be alleviated. \emph{Theory Into Practice}, \emph{55}(2), 104--111.

>>>>

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={New approaches to teachers' experience of stress: Do heart rate measurements with fitness trackers provide an efficient, inexpensive, and robust measurement method?},
  pdfauthor={Mandy Klatt1, Peer Keßler1, Gregor Kachel1,2, Christin Lotz1, \& Anne Deiglmayr1},
  pdflang={en-EN},
  pdfkeywords={heart rate; photoplethysmography; wearable electronic device; expertise differences},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{New approaches to teachers' experience of stress: Do heart rate measurements with fitness trackers provide an efficient, inexpensive, and robust measurement method?}
\author{Mandy Klatt\textsuperscript{1}, Peer Keßler\textsuperscript{1}, Gregor Kachel\textsuperscript{1,2}, Christin Lotz\textsuperscript{1}, \& Anne Deiglmayr\textsuperscript{1}}
\date{}

@@

The authors made the following contributions. Mandy Klatt: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Peer Keßler: Writing - Original Draft Preparation, Writing - Review \& Editing; Gregor Kachel: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Christin Lotz: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Anne Deiglmayr: Supervision.

@@

Prospective teachers experience the first five years as a particular professional challenge, as they are associated with a high workload (e.g., class size, poor student behavior) and resulting stress in the teaching profession (Struyven \& Vanthournout, 2014).

Psychosomatic stress symptoms, such as increasing heart rate, result from an interaction of situational stressors and person-specific available resources {[}``transactional stress model''; Lazarus and Folkman (1987); Obbarius, Fischer, Liegl, Obbarius, and Rose (2021){]}. If teachers are exposed to a stressful teaching-learning environment for a long time and do not have the sufficient resources and coping strategies, this can lead to burnout.

Fisher (2011) showed in her study of 400 secondary school teachers that burnout scores differed significantly between novice and experienced teachers, with novices having higher burnout scores. In addition, she found that a low number of years of professional experience was a statistically significant predictor of stress in the teaching profession, along with job satisfaction and burnout.

In addition to self-reported indicators of stress (e.g., emotional exhaustion), psychophysiological indicators (e.g., heart rate) are becoming increasingly important in research. Heart rate as the number of heart contractions per minute (Löllgen, 2015) is an important indicator of physical or emotional stress. Increased workload is associated with increased heart rate (Sachs, 2014). Arguably, physiological data provide a more objective and non-reactive measure of stress in the classroom (Runge et al., 2020). However, in an educational context, this requires the use of low-cost, non-invasive instruments. Fitness trackers worn on the wrist could fill this research gap (Ferguson, Rowlands, Olds, \& Maher, 2015).

@@

\leavevmode\vadjust pre{\hypertarget{ref-ferguson2015validity}{}}%
Ferguson, T., Rowlands, A. V., Olds, T., \& Maher, C. (2015). The validity of consumer-level, activity monitors in healthy adults worn in free-living conditions: A cross-sectional study. \emph{International Journal of Behavioral Nutrition and Physical Activity}, \emph{12}(1), 1--9.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lazarus1987transactional}{}}%
Lazarus, R. S., \& Folkman, S. (1987). Transactional theory and research on emotions and coping. \emph{European Journal of Personality}, \emph{1}(3), 141--169.

\leavevmode\vadjust pre{\hypertarget{ref-Luxf6llgen2015}{}}%
Löllgen, H. (2015). Herzfrequenz und blutdruck. In J. Niebauer (Ed.), \emph{Sportkardiologie} (pp. 87--105). Berlin, Heidelberg: Springer Berlin Heidelberg. \url{https://doi.org/10.1007/978-3-662-43711-7_9}

\leavevmode\vadjust pre{\hypertarget{ref-obbarius2021modified}{}}%
Obbarius, N., Fischer, F., Liegl, G., Obbarius, A., \& Rose, M. (2021). A modified version of the transactional stress concept according to lazarus and folkman was confirmed in a psychosomatic inpatient sample. \emph{Frontiers in Psychology}, \emph{12}, 584333.

@@
